---
title: "NHANES Dataset Complex Table Report"
author: "Giba Barrie"
date: "2026-10-05"
output: pdf_document
bibliography: references.bib
---

```{r setup, include=FALSE}
library(tidyverse)
library(rio)
library(here)
library(knitr)
library(kableExtra)

cleaned_NHANES <- import(here("data", "cleaned_NHANES.csv"))
colnames(cleaned_NHANES) <- tolower(colnames(cleaned_NHANES))
cleaned_NHANES <- cleaned_NHANES %>% 
  mutate(
    gender = tolower(gender),
    ethnicity_2 = tolower(ethnicity_2)
  )

# Calculate the average systolic blood pressure across the 4 columns
cleaned_NHANES <- cleaned_NHANES %>% 
  mutate(bpsysave = rowMeans(select(., systolic_bp_1, systolic_bp_2, systolic_bp_3, systolic_bp_4), na.rm = TRUE))
```

# Introduction
The National Health and Nutrition Examination Survey (NHANES) is a major program of studies designed to assess the health and nutritional status of adults and children in the United States [@nhanes]. It  provides an overview of various variables including gender, ethnicity, and blood pressure. 

```{r create-bib, echo=FALSE, message=FALSE, warning=FALSE}
# Bypasses the cat engine bug by writing the file natively
bib_content <- c(
  "@misc{nhanes,",
  "  author = {{Centers for Disease Control and Prevention}},",
  "  title = {National Health and Nutrition Examination Survey},",
  "  year = {2026},",
  "  url = {https://cdc.gov}",
  "}"
)
writeLines(bib_content, "references.bib")
```

# Demographic & Blood Pressure Metrics by Wave

```{r table1, echo=FALSE}
# Aggregate sample size, male counts, and blood pressure stats by wave
wave_summary <- cleaned_NHANES %>%
  group_by(wave_id) %>%
  summarise(
    Total_N = n(),
    Male_N = sum(gender == "male", na.rm = TRUE),
    Male_Pct = round((Male_N / Total_N) * 100, 1),
    BP_Min = round(min(bpsysave, na.rm = TRUE), 1),
    BP_Max = round(max(bpsysave, na.rm = TRUE), 1),
    BP_Med = round(median(bpsysave, na.rm = TRUE), 1),
    BP_Mean = round(mean(bpsysave, na.rm = TRUE), 1),
    BP_SD = round(sd(bpsysave, na.rm = TRUE), 1),
    .groups = "drop"
  )

kable(wave_summary, 
      format = "latex",
      booktabs = TRUE,
      col.names = c("Wave ID", "Total N", "Count", "%", "Min", "Max", "Median", "Mean", "SD"),
      caption = "Sample Overview and Systolic Blood Pressure Statistics by Wave") %>%
  kable_styling(latex_options = c("striped", "hold_position")) %>%
  add_header_above(c(" " = 2, "Male Demographics" = 2, "Average Systolic BP (mm Hg)" = 5)) %>%
  row_spec(0, bold = TRUE, background = "lightgray")
```

# Ethnic Distribution by Study Wave
To maintain layout legibility within the PDF borders, the frequency counts and calculated percentages for participant ethnic categories are detailed below.

```{r table2, echo=FALSE}
ethnicity_summary <- cleaned_NHANES %>%
  group_by(wave_id, ethnicity_2) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(pct = round((count / sum(count)) * 100, 1)) %>%
  mutate(display_val = paste0(count, " (", pct, "%)")) %>%
  select(wave_id, ethnicity_2, display_val) %>%
  pivot_wider(names_from = ethnicity_2, values_from = display_val, values_fill = "0 (0%)")

# Render the secondary table
kable(ethnicity_summary, 
      format = "latex",
      booktabs = TRUE,
      caption = "Participant Ethnicity Distribution Counts and Percentages by Wave") %>%
  kable_styling(latex_options = c("striped", "hold_position")) %>%
  row_spec(0, bold = TRUE, background = "lightgray")
```

# References
