---
title: "Week 3 Assignment"
author: "Giba Barrie"
date: "2026-09-27"
output: html_document
--- 

```{r lib}
library(tidyverse)
library(rio)
library(here)
library(ggpubr)
library(cowplot)
```

##Exercise 1
#Recreating figure 1 - Age distribution of participants by Gender

```{r histogram}
cleaned_NHANES <- import(here("data", "cleaned_NHANES.csv"))

ggplot(data = cleaned_NHANES, aes(x = age, fill = gender)) +
  geom_histogram(binwidth = 5, position = "dodge") + 
  labs(
    title = "Age distribution of participants by Gender",
    x = "Age (Years)",
    y = "Count",
    fill = "Gender"
    )
```
#Recreating figure 2
```{r bargraph}
ggplot(data = cleaned_NHANES, aes(x = ethnicity_2, fill = gender)) +
  geom_bar(position = "fill") + 
  labs(
    title = "Gender proportion by ethnicity",
    x = "Ethnicity",
    fill = "Gender"
  )
```

#Combining two figures 
```{r setup} 
p1 <- ggplot(data = cleaned_NHANES, aes(x = age, fill = gender)) +
  geom_histogram(binwidth = 5, position = "dodge") + 
  labs(
    title = "Age distribution of participants by Gender",
    x = "Age (Years)",
    y = "Count",
    fill = "Gender"
  )

p2 <- ggplot(data = cleaned_NHANES, aes(x = ethnicity_2, fill = gender)) +
  geom_bar(position = "fill") + 
  labs(
    title = "Gender proportion by ethnicity",
    x = "Age (Years)",
    y = "Count",
    fill = "Gender")
```

#Using cowplot 
```{r combined w cowplot}
 combined <- plot_grid (p1, p2, labels = c( "A", "B"), ncol = 2 )
print(combined)
```

#Combining using ggarrange 
In this instance, I believe plot_grid worked best aesthetically because it aligned the graphs better. 

```{r combined w ggarrange}
combined <- ggarrange(p1, p2, 
                      labels = c("A", "B"), 
                      ncol = 2,
                      nrow =1, 
                      common.legend = TRUE, legend = "right")
print(combined)
```
##Exercise 2 
In comparing the two ethnicity variables, I believe ethnicity2 is the better option as it has more racial categories. 

For the age figure, the highest distribiution is among those aged 2. 
For the gender figure, females are more representated than males. 
For the Ethnicity 2 figure, white people account for the highest racial group. 
```{r exercise 2-graphs, message=FALSE, warning=FALSE}
library(ggplot2)
#Age figure 
figure_age <- ggplot(data = cleaned_NHANES, aes(x = age)) +
  geom_histogram(color = "black") + 
  labs(
    title = "Distribution of Participant Age",
    x = "Age (Years)",
    y = "Count"
  )
print(figure_age)
ggsave("figure_age.png", plot = figure_age, width = 6, height = 4)

#Gender figure 
gender_figure <- ggplot(data = cleaned_NHANES, aes(x = gender, fill = gender)) +
  geom_bar() + 
  labs(
    title = "Distribution of Participant Gender",
    x = "Gender",
    y = "Count") + theme(legend.position = "none")
print(gender_figure)
ggsave("gender_figure.png", plot = gender_figure, width = 6, height = 4)

#Ethnicity variable 1
ethnicity1 <- ggplot(data = cleaned_NHANES, aes(x = ethnicity_1, fill = ethnicity_1)) +
  geom_bar() + 
  labs(
    title = "Distribution of Ethnicity (1)",
    x = "Ethnicity Category",
    y = "Count"
  )
print(ethnicity1)
ggsave("ethnicity1.png", plot = ethnicity1, width = 6, height = 4)


#Ethnicity variable 2 
ethnicity2 <- ggplot(data = cleaned_NHANES, aes(x = ethnicity_2, fill = ethnicity_2)) +
  geom_bar() + 
  labs(
    title = "Distribution of Ethnicity (2)",
    x = "Ethnicity Category",
    y = "Count"
  )
print(ethnicity2)
ggsave("ethnicity2.png", plot = ethnicity2, width = 6, height = 4)
```

##Exercise 3 
From a first glance, the graph is overwhelming as it is cluttered. The colours look too similar, and the legend not being labelled is confusing. 
I would use a color scheme that is easier to follow and a legend that is labelled clearly 

```{r exercise 3-graphs, message=FALSE, warning=FALSE}

diet_data <- import(here("data", "diet.csv"))
baseline_weights <- diet_data %>% 
  filter(Week == 0) %>% 
  select(Participant, baseline = Weight)

improved_data <- diet_data %>% 
  left_join(baseline_weights, by = "Participant") %>% 
  mutate(weight_change = Weight - baseline)

ggplot(improved_data, aes(x = Week, y = weight_change)) +
  geom_smooth(color = "pink")+
  geom_point() +
  theme_minimal() +
  labs(
    title = "Participant weight change with diet", 
    x = "Week",
    y = "Weight change from week 0"
  )
```

##Exercise 4 

```{r exercise 4-graphs, message=FALSE, warning=FALSE}

nhanes_data <- cleaned_NHANES %>% 
  mutate(
    average_sbp = rowMeans(
      cbind(systolic_bp_1, systolic_bp_2, systolic_bp_3, systolic_bp_4), 
      na.rm = TRUE
    )
  )

nhanes_filtered <- nhanes_data %>% 
  filter(
    age == age,
    gender == gender,
    average_sbp == average_sbp
  )

nhanes_final <- nhanes_filtered %>% 
  mutate(
    sbp_category = case_when(
      average_sbp < 120 ~ "Normal",
      average_sbp >= 120 & average_sbp < 130 ~ "Elevated",
      average_sbp >= 130 & average_sbp < 140 ~ "Stage 1 hypertension",
      average_sbp >= 140 ~ "Stage 2 hypertension"
    ),
    sbp_category = factor(
      sbp_category, 
      levels = c("Normal", "Elevated", "Stage 1 hypertension", "Stage 2 hypertension")
    )
  )

ggplot(nhanes_final, aes(x = sbp_category)) +
  geom_bar(fill = "purple") +
  theme_minimal() +
  labs(
    title = "Distribution of Blood Pressure Categories",
    x = "Hypertension Category",
    y = "Count of Individuals"
  )

ggplot(nhanes_final, aes(x = gender)) +
  geom_bar(fill = "lightblue") +
  theme_minimal() +
  labs(
    title = "Gender Distribution in Final Sample",
    x = "Gender",
    y = "Count of Individuals"
  )

# In my sample here versus my previous figure, the sample size reduced significantly. In particular, females had a count of about 15000, and now the count of individuals is about 9500, leading to a change of approximately 5500 females. 

#Final visualization 

ggplot(nhanes_final, aes(x = sbp_category)) +
  geom_bar(fill = "darkred") +
  facet_grid(gender ~ age_cat) +
    scale_fill_brewer(palette = "Red") + 
  theme_minimal() +
  labs(
    title = "Hypertension Distributions Across Gender and Age Groups",
    x = "Systolic Blood Pressure Category",
    y = "Count" 
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

